---
title: "RNA_seq_definitive_edition_17_03_22"
author: "Antonin COLAJANNI"
date: "17/03/2022"
output: html_document
---


# Analyse des données RNAseq DepMap (CCLE ; cancer) et des 57 epigenomes de Roadmap

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r import, echo=T, results='hide', message=F, warning=F}
# W O R K I N G  D I R E C T O R Y
main.dir = "/shared/projects/chrom_enhancer_bc"
setwd(main.dir)

data.dir = file.path(main.dir,"data/expression/rdata")


# I M P O R T
library(pheatmap  )
library(ggpubr)
library(sva)
library(edgeR)
library(DESeq2)
library(DT)
library(stringr)
library(dplyr)
library(reshape2)
library(FactoMineR)
library(factoextra)
library(grid)
source("./scripts/Normalization.R")
load(file.path(data.dir,"cancer_healthy_expression.RData"))


# F U N C T I O N S

# Centering by the median
MedianCentering <- function(x){
  (x - median(x))
}

get_color_scale_pheatmap = function(matrix,ncolors = 50,threshold = 1){
  
  range <- max(abs(matrix))
  myBreaks = seq(-range, range, length.out = ncolors)
  myBreaks[myBreaks > -threshold  &  myBreaks < threshold ] = 0
  myBreaks = unique(myBreaks)
  
  paletteLength <- length(myBreaks)
  myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength-1)
  return(list("colorscale" = myColor, "breaks" = myBreaks)) }

# D A T A   P R O C E S S 

# change format
cell_lines = names(expr)[! names(expr) %in% c("gene_name","ENSEMBL_id")]
expr = select(expr, -gene_name)
# as numeric
expr <- cbind("ENSEMBL_id"=expr$ENSEMBL_id, 
              mutate_all(select(expr,all_of(cell_lines)), function(x) as.numeric(as.character(x))))

# Remove lowly expressed genes + 1 duplicated id
expr <- expr %>%
  filter(! duplicated(ENSEMBL_id))

row.names(expr) = expr$ENSEMBL_id
expr$ENSEMBL_id = NULL

Row_sums = rowSums(expr)
filtered_expr = expr[Row_sums > 3*ncol(expr),]
```

```{r print_table}

DT::datatable(head(filtered_expr))
```

```{r batch_limma}
# Parameters
design = c(rep("CCLE", 20 ), rep("RoadMap",length(cell_lines)-20))
count.matrix = data.matrix(filtered_expr)



All_normalisations = function(count.matrix,design){
  norm_list = list()
  tools=c("TMM", "TMMwsp", "RLE", "upperquartile")
  # edgeR
  for (tool in tools){
    message(tool)
    edgeR.dgelist = DGEList(counts = count.matrix, group = factor(design))
    nf = calcNormFactors(edgeR.dgelist, method=tool)
    edgeR_norm <- cpm(nf, normalized.lib.sizes=TRUE, log = TRUE)
    norm_list[[tool]] = edgeR_norm  
  }
  #DESeq2
  message("DESeq2")
  deseq_norm = tools.norm.RNAseq(count.matrix, tool = "vst2", design = design )
  norm_list[["DESeq2"]] = deseq_norm
  return(norm_list)
}


norm_list = All_normalisations(count.matrix,design)


## Applying limma batch correction
norm_list_limma = lapply(norm_list, function(x) removeBatchEffect(x, design) )
norm_list_limma_melted = lapply(norm_list_limma, function(x) reshape2::melt(x) )


```


```{r batch_combat}
# ComBat batch corrections : needs raw counts
load(file.path(data.dir,"ComBat_seq_correction.RData"))

norm_list_ComBat = All_normalisations(adjusted,design)
norm_list_ComBat_melted = lapply(norm_list_ComBat, function(x) reshape2::melt(x) )



```

### Scatterplot Limma batch correction vs ComBat_seq

```{r scatter0}

k562 = colnames(norm_list_limma[['TMM']])[grepl("K562",colnames(norm_list_limma[['TMM']]))] 


Limma_k562 = lapply(norm_list_limma, function(x) select(data.frame(x), all_of(k562) ))
ComBat_k562 =  lapply(norm_list_ComBat, function(x) select(data.frame(x), all_of(k562) ))


polr3g ="ENSG00000113356"
polr3g_full_name = rownames(count.matrix)[grepl(pattern = polr3g, rownames(count.matrix))]




ggplot(select(data.frame(log2(count.matrix+1)), K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE,K562), aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             data.frame(log2(count.matrix+1))[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 20, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 17.5, aes(label = ..rr.label..)) +
  ggtitle("Raw counts (log2+1)")


```


```{r scatter1}
#K562



ggplot(Limma_k562$TMM, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             Limma_k562$TMM[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("TMM normalisation - Limma batch correction")

```

```{r scatter2}


ggplot(Limma_k562$TMMwsp, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             Limma_k562$TMMwsp[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("TMMwsp normalisation - Limma batch correction")

```

```{r scatter3}


ggplot(Limma_k562$RLE, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             Limma_k562$RLE[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("RLE normalisation - Limma batch correction")

```

```{r scatter4}


ggplot(Limma_k562$upperquartile, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             Limma_k562$upperquartile[polr3g_full_name,], 
             color = "red",
             size = 3)+
  
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("Upperquartile normalisation - Limma batch correction")

```

```{r scatter5}


ggplot(Limma_k562$DESeq2, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             Limma_k562$DESeq2[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 20, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 15, aes(label = ..rr.label..)) +
  ggtitle("vst normalisation (DESeq2) - Limma batch correction")

```


```{r scatter12}

ggplot(ComBat_k562$TMM, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             ComBat_k562$TMM[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("TMM normalisation - Combat_seq correction")

```

```{r scatter22}


ggplot(ComBat_k562$TMMwsp, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             ComBat_k562$TMMwsp[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("TMMwsp normalisation - Combat_seq correction ")

```

```{r scatter32}


ggplot(ComBat_k562$RLE, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             ComBat_k562$RLE[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("RLE normalisation - Combat_seq correction")

```

```{r scatter42}


ggplot(ComBat_k562$upperquartile, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() + 
  geom_point(data = 
             ComBat_k562$upperquartile[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 15, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 10, aes(label = ..rr.label..)) +
  ggtitle("Upperquartile normalisation - Combat_seq correction")

```

```{r scatter52}


ggplot(ComBat_k562$DESeq2, aes(x=K562_HAEMATOPOIETIC_AND_LYMPHOID_TISSUE, y=K562)) + 
  geom_point() +
  geom_point(data = 
             ComBat_k562$DESeq2[polr3g_full_name,], 
             color = "red",
             size = 3)+
  geom_smooth(method = "lm") +
  stat_regline_equation(label.y = 20, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 18, aes(label = ..rr.label..)) +
  ggtitle("vst normalisation (DESeq2) - Combat_seq correction")
```




# Test des normalisations seules vs log2(raw+1)
```{r boxplots_norm ,fig.width=15, fig.height=10}

ctcfl = "ENSG00000124092"
ctcf =  "ENSG00000102974"
pou5f1 ="ENSG00000204531"
myc =   "ENSG00000136997"
polr3gl="ENSG00000121851"
polr3g ="ENSG00000113356"

CCLE = cell_lines[1:20]
ROADMAP = cell_lines[21:length(names(expr))]
#melted_expr = reshape2::melt(edgeR_norm)

condition = ifelse(norm_list_ComBat_melted$TMM$Var2 %in% CCLE, yes = "CCLE", no = "RoadMap")

# cancer lines of interest : 
of_interest = c("MDAMB231","MDAMB436","HCC1937", "HCC1806",
                "HCC1954","MDAMB468","K562")
color_cell = ifelse(grepl(paste(of_interest,collapse = "|"),  cell_lines ), yes = "red", no ="black")

expression_plot = function(melted_df){
plot = ggplot(data = melted_df, aes(x=Var2, y=value, color = condition)) + 
  geom_boxplot()   +
  geom_text(aes(x=Var2,
                label=ifelse(grepl(polr3g,Var1),".",''))
            , size= 10, color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = color_cell),
          axis.text=element_text(size=6))  +
  xlab("cell_lines") + ylab("normalized_expression")
  return(plot)
}

compute_plot = function(liste, title = "Normalisation"){
  first_part_title = "Expression boxplot :"
  c = 1
  plot_list = list()
  for (df in liste) {
    tmp_title = paste(first_part_title, names(liste)[c] ,title)
    plot = expression_plot(df) + ggtitle(tmp_title)
    plot_list[[ names(liste)[c] ]] = plot
  
    c = c+1
  }
  return(plot_list) }



# Raw + log2
melt_raw = reshape2::melt(log2(count.matrix+1))
raw_log = expression_plot(melt_raw) + ggtitle("Expression boxplot : Raw expression (log2+1)")


norm_list_melted = lapply(norm_list, function(x) reshape2::melt(x) )
norm_plot = compute_plot(norm_list_melted)
print(norm_plot$DESeq2)


for (plot in norm_plot){
  print(plot)
}


```

# Test des batchs correction seules vs  DESeq2 norm vs log2(raw+1)


```{r boxplots_norm2,fig.width=15, fig.height=10}
# RAW
raw_log

# Norm + batch(limma)
batch_norm_melt = norm_list_limma_melted$DESeq2
expression_plot(batch_norm_melt) + ggtitle("Expression boxplot : Normalized expression + batch correction (limma)")

# Norm + batch(ComBat)
batch_norm_combat_melt =  norm_list_ComBat_melted$DESeq2
expression_plot(batch_norm_combat_melt) + ggtitle("Expression boxplot : Normalized expression + batch correction (Combat)")

# Raw + batch(ComBat)
batch_raw_combat_melt = reshape2::melt(log2(adjusted+1))
expression_plot(batch_raw_combat_melt) + ggtitle("Expression boxplot : Raw expression (log2+1) + batch correction (Combat)")


# raw + batch(limma)
batch_raw = removeBatchEffect(log2(count.matrix+1), design)
batch_raw_melt = reshape2::melt(batch_raw)
expression_plot(batch_raw_melt) + ggtitle("Expression boxplot : Raw expression (log2+1) + batch correction (limma)")


```




# Vérification avec PCA
## Raw
```{r PCA1, fig.width=10, fig.height=10}

raw_matrix = data.frame(log2(count.matrix+1))


extract_top_varying_genes = function(matrix, threshold=0.85){
  IQRs = data.frame("IQR" = apply(matrix, 1, IQR))
  IQRs$IQR = as.numeric(IQRs$IQR)
  return( rownames(filter(IQRs, IQRs$IQR> quantile(IQRs$IQR,threshold)))  )
}


pca_subset = raw_matrix[extract_top_varying_genes(raw_matrix) , ]


```

```{r PCAraw, fig.width=15, fig.height=15}

of_interest = c("MDAMB231","MDAMB436","HCC1937", "HCC1806",
                "HCC1954","MDAMB468")

design_PCA = ifelse(grepl("K562", cell_lines), yes = "K562", no = design)
design_PCA = ifelse(grepl(paste(of_interest,collapse = "|"), cell_lines), yes = "TNBC", no = design_PCA)

res.pca = PCA(t(pca_subset), graph = F)
fviz_pca_ind(res.pca, 
             col.ind = design_PCA,
             repel = TRUE,
             palette = "Set1",
             ggrepel.max.overlaps = 1) + labs(title ="PCA raw matrix",)



```

## Normalisation

```{r PCANorm, fig.width=15, fig.height=15}

norm_pca = data.frame(t(norm_list$DESeq2))
norm_pca = norm_pca[,extract_top_varying_genes(norm_list$DESeq2)]


res.pca = PCA(norm_pca,  graph = F)
fviz_pca_ind(res.pca, 
             col.ind = design_PCA,
             repel = TRUE,
             palette = "Set1",
             ggrepel.max.overlaps = 1) + labs(title ="PCA normalisad count (DESeq2)",)





```

## Normalisation + Batch correction (ComBat_seq)
```{r PCANorm2, fig.width=15, fig.height=15}

norm_pca_combat = data.frame(t(norm_list_ComBat$DESeq2))
norm_pca_combat = norm_pca_combat[,extract_top_varying_genes(norm_list_ComBat$DESeq2,0.85)]

res.pca = PCA(norm_pca_combat,  graph = F)
fviz_pca_ind(res.pca, 
             col.ind = design_PCA,
             repel = TRUE,
             palette = "Set1",
             ggrepel.max.overlaps = 1) + labs(title ="PCA normalisad count (DESeq2) + ComBat_seq",)





```


## Normalisation + Batch correction (limma)
```{r PCANorm3, fig.width=15, fig.height=15}

#norm_pca_limma = data.frame(t(norm_list_limma$DESeq2[rownames(filter(IQRs, IQRs$IQR>7)) , ]))

norm_pca_limma = data.frame(t(norm_list_limma$DESeq2))
norm_pca_limma = norm_pca_limma[, extract_top_varying_genes(norm_list_limma$DESeq2,0.85) ]

res.pca = PCA(norm_pca_limma,  graph = F)
fviz_pca_ind(res.pca, 
             col.ind = design_PCA,
             repel = TRUE,
             palette = "Set1",
             ggrepel.max.overlaps = 1) + labs(title ="PCA normalisad count (DESeq2) + limma ",)





```




# Heatmap d'expression de la région autour de POLR3G


```{r heatmap, fig.width=15, fig.height=10}

ComBat_norm = norm_list_ComBat_melted$DESeq2

# VARIABLES
TNBC = c("MDAMB231","MDAMB436","HCC1937", "HCC1806",
         "HCC1954","MDAMB468")

of_interest = c(
  "POU5F1" ="ENSG00000204531",
  "MYC" =   "ENSG00000136997",
  "POLR3GL"="ENSG00000121851",
  "POLR3G" ="ENSG00000113356",
  "MEF2C" = "ENSG00000081189",
  "CETN3" = "ENSG00000153140",
  "MBLAC2" ="ENSG00000176055",
  "LYSMD3" ="ENSG00000176018",
  "ADGRV1" ="ENSG00000164199")
of_interest = data_frame("ENS_id" = of_interest, "gene_name" = names(of_interest))

# Filtering matrix with only gene of interests
tmp = filter(ComBat_norm, grepl(paste(of_interest$ENS_id,collapse="|"),ComBat_norm$Var1))
tmp$copie = substr(tmp$Var1,1,15)


# merge dataframes to correctly match ENS_ID with gene name
to_heatmap = select(merge(tmp,of_interest, by.x = "copie", by.y = "ENS_id"),
                    gene_name,Var2 ,value)
colnames(to_heatmap) = c("gene_name","cell_line","expression")

# melt ==> classical matrix
heat = reshape2::dcast(to_heatmap, gene_name~cell_line)
row.names(heat) = heat$gene_name
heat$gene_name = NULL



# Median centering
heat_norm = t(apply(heat, 1, MedianCentering))

# Create color scale
scale = get_color_scale_pheatmap(heat_norm,ncolors = 50,threshold = 0.75)


annotation_heatmap = data.frame("data_source" = design, row.names = cell_lines)
color = c("red","blue")
names(color) = c("CCLE","RoadMap")


annotation_heatmap$state = ifelse(
  grepl(paste(TNBC,collapse="|") ,row.names(annotation_heatmap)), 
  yes = "TNBC", no = "other")



ComBat_heatmap = data.frame(t(heat_norm))
ComBat_heatmap$colors = ifelse(grepl(paste(TNBC,collapse="|") ,row.names(ComBat_heatmap)), 
                                                  yes = "red", no = "darkgrey")
ComBat_heatmap = ComBat_heatmap[order(-ComBat_heatmap$POLR3G) , ]

ComBat_heatmap = ComBat_heatmap[,c("CETN3","MBLAC2","LYSMD3","MYC","POLR3GL","ADGRV1","MEF2C","POLR3G","colors")]


map = pheatmap(select(ComBat_heatmap,-colors),
         color = scale$colorscale,
         breaks = scale$breaks,
         cluster_rows = F, cluster_cols = F,
         angle_col = 0,
         fontsize_row = 5,
         main = "Expression heatmap : ComBat batch correction ")


#tmp = ComBat_heatmap
map$gtable$grobs[[4]]$gp=gpar(col= ComBat_heatmap$colors, fontsize = 5)


print(map) 


```

```{r heatmap2, fig.width=15, fig.height=10}

limma_norm = norm_list_limma_melted$DESeq2


tmp = filter(limma_norm, grepl(paste(of_interest$ENS_id,collapse="|"),limma_norm$Var1))
tmp$copie = substr(tmp$Var1,1,15)



to_heatmap = select(merge(tmp,of_interest, by.x = "copie", by.y = "ENS_id"),
                    gene_name,Var2 ,value)
colnames(to_heatmap) = c("gene_name","cell_line","expression")

heat = reshape2::dcast(to_heatmap, gene_name~cell_line)
row.names(heat) = heat$gene_name
heat$gene_name = NULL


heat_norm = t(apply(heat, 1, MedianCentering))
scale = get_color_scale_pheatmap(heat_norm,ncolors = 50,threshold = 0.75)


annotation_heatmap = data.frame("data_source" = design, row.names = cell_lines)
color = c("red","blue")
names(color) = c("CCLE","RoadMap")


annotation_heatmap$state = ifelse(
  grepl(paste(TNBC,collapse="|") ,row.names(annotation_heatmap)), 
  yes = "TNBC", no = "other")



limma_heatmap = data.frame(t(heat_norm))
limma_heatmap$colors = ifelse(grepl(paste(TNBC,collapse="|") ,row.names(limma_heatmap)), 
                                                  yes = "red", no = "darkgrey")

limma_heatmap = limma_heatmap[order(-limma_heatmap$POLR3G) , ]
limma_heatmap = limma_heatmap[,c("CETN3","MBLAC2","LYSMD3","MYC","POLR3GL","ADGRV1","MEF2C","POLR3G","colors")]


map = pheatmap(select(limma_heatmap,-colors),
         color = scale$colorscale,
         breaks = scale$breaks,
         cluster_rows = F, cluster_cols = F,
         angle_col = 0,
         fontsize_row = 5,
         main = "Expression heatmap : limma batch correction ")


#tmp = ComBat_heatmap
map$gtable$grobs[[4]]$gp=gpar(col= limma_heatmap$colors, fontsize = 5)
print(map) 


```


```{r ggpairs, fig.width=15, fig.height=10}
library(GGally)

ComBat_norm = data.frame(norm_list_ComBat$DESeq2)

tmp = filter(ComBat_norm, grepl(paste(of_interest$ENS_id,collapse="|"),substr(row.names(ComBat_norm),1,15) ))
tmp$ENS_id = substr(rownames(tmp),1,15)
tmp = merge(tmp, of_interest, by = "ENS_id")
row.names(tmp) = tmp$gene_name
to_be_plotted = select(tmp, -gene_name, -ENS_id)


design = c(rep("CCLE", 20 ), rep("RoadMap",length(cell_lines)-20))
#to_be_plotted = rbind(to_be_plotted, "condition" = design)

#to_be_plotted

ggpairs(as.data.frame(t(to_be_plotted)),
        #aes(colors=design)
        mapping = ggplot2::aes(color =design))+
  theme_minimal()

```


```{r ggpairs2, fig.width=15, fig.height=10}


limma_norm = data.frame(norm_list_limma$DESeq2)

tmp = filter(limma_norm, grepl(paste(of_interest$ENS_id,collapse="|"),substr(row.names(limma_norm),1,15) ))
tmp$ENS_id = substr(rownames(tmp),1,15)
tmp = merge(tmp, of_interest, by = "ENS_id")
row.names(tmp) = tmp$gene_name
to_be_plotted = select(tmp, -gene_name, -ENS_id)


design = c(rep("CCLE", 20 ), rep("RoadMap",length(cell_lines)-20))
#to_be_plotted = rbind(to_be_plotted, "condition" = design)

#to_be_plotted

ggpairs(as.data.frame(t(to_be_plotted)),
        #aes(colors=design)
        mapping = ggplot2::aes(color =design))+
  theme_minimal()

```