---
title: "interactions_polr_3g_15_02_22"
author: "Antonin_Colajanni"
date: "15/02/2022"
output: html_document
---

```{r setup, error=FALSE, warning=FALSE, message=FALSE}
# W O R K I N G  D I R E C T O R Y
main.dir = "/shared/projects/chrom_enhancer_bc"
setwd(main.dir)

data.dir = file.path(main.dir,"data/interaction/rdata/")


# I M P O R T
source("./scripts/fun_interactions.r")
load(file.path(data.dir,"/BENGI_benchmark_pchic.RData"))
load(file.path(data.dir,"/3div_0.05.Rdata"))
load(file.path(data.dir,"/raw_BENGI_interactions_POLR3G.RData"))
library(UpSetR)

data = File_list_filtered0.05
rm(File_list_filtered0.05)
data_05 = lapply(data, hic_format )
data_01 = lapply(data_05, function(x) filter_pvalue(df=x, seuil = 2))


polr3g_exact = GRanges(
  seqnames = Rle("chr5"),
  IRanges(89768410, 89777404))
```

```{r preupset05}
# interactions lists 
interactions_05 = lapply(data_05, function(x) subset_by_overlap(ROI = polr3g_exact, df = x))
interaction_polr3g_05 = lapply(interactions_05, function(x) x$interaction_polr3g)
# as dataframe
interactions_df_05 = ldply(interactions_05)
# listing of Cell_lines
cell_lines = unique(interactions_df_05$.id)
```

## Upset plot p<0.05

```{r upset, fig.width=8, fig.height=5}
upset(fromList(interaction_polr3g_05),
      sets = cell_lines, sets.bar.color = "#56B4E9",
      order.by = "freq", mb.ratio = c(0.4,0.6),
      empty.intersections = NULL, set_size.show = TRUE,
      set_size.scale_max = 50)
```


```{r preupset01}
# interactions lists 
interactions_01 = lapply(data_01, function(x) subset_by_overlap(ROI = polr3g_exact, df = x))
interaction_polr3g_01 = lapply(interactions_01, function(x) x$interaction_polr3g)
# as dataframe
interactions_df_01 = ldply(interactions_01)
# listing of Cell_lines
cell_lines = unique(interactions_df_01$.id)
```


## Upset plot p<0.01
```{r upset012, fig.width=8, fig.height=5}
upset(fromList(interaction_polr3g_01),
      sets = cell_lines, sets.bar.color = "#56B4E9",
      order.by = c("degree"), mb.ratio = c(0.4,0.6),
      empty.intersections = NULL, set_size.show = TRUE,
      set_size.scale_max = 20)
```

# Stacked Barplot 

## p<0.05
```{r stack15, fig.width=8, fig.height=5}
# All the interactions that are shared 
freq_05 = data.frame(table(interactions_df_05$interaction_polr3g))
unique = as.vector(freq_05[freq_05$Freq == 1,]$Var1)
not_unique = as.vector(freq_05[freq_05$Freq >= 2,]$Var1)

# counting interactions
find_unique <- function(df, unique){
  tmp = ifelse(df$interaction_polr3g %in% unique, yes = TRUE, no = FALSE)
  return(data.frame("interactions" = c(sum(tmp==TRUE), sum(tmp==FALSE)),
                    "condition" = c("unique", "common"))) }

interactions_unique = lapply(interactions_05, 
                             function(x) find_unique(x, unique))

interactions_unique = ldply(interactions_unique)

```

```{r stack152, fig.width=10, fig.height=8}
p <- ggplot(data = interactions_unique, aes(x = reorder(.id,interactions), y = interactions)) +
  geom_col(aes(fill = condition), width = 0.7)+
  geom_text(data=subset(interactions_unique, interactions!=0),
    position = position_stack(vjust = 0.5),
    size = 4,
    aes(y = , label = interactions, group =condition), color = "white")+
    scale_fill_manual(values = c("#00539CFF", "orangered"))+
    theme_classic() + coord_flip() +  
    xlab("Cell line") + ylab("Contact number") + 
    ggtitle("Number of statistically significant interactions with POLR3G (p<0.05)") 
p
```

## Pie chart
```{r pie_chart_prep, fig.width=4, fig.height=4}
pie_df = data.frame("value" = c(length(unique), length(not_unique)),
                    "condition" = c("unique","common"))
```

```{r piechart, fig.width=5, fig.height=4}
ggplot(pie_df, aes(x="", y=value, fill=condition)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  scale_fill_manual(values = c("#00539CFF", "orangered")) + 
  geom_text(
    position = position_stack(vjust = 0.5), 
    size = 5, 
    aes(y = ,label = value, group =condition), color = "white")+
    ggtitle("Number of statistically significant interactions \n with POLR3G (p<0.05)") 

```


# p<0.01

```{r stack153, fig.width=8, fig.height=5}
# All the interactions that are shared 
freq_01 = data.frame(table(interactions_df_01$interaction_polr3g))
unique = as.vector(freq_01[freq_01$Freq == 1,]$Var1)
not_unique = as.vector(freq_01[freq_01$Freq >= 2,]$Var1)

# counting interactions
find_unique <- function(df, unique){
  tmp = ifelse(df$interaction_polr3g %in% unique, yes = TRUE, no = FALSE)
  return(data.frame("interactions" = c(sum(tmp==TRUE), sum(tmp==FALSE)),
                    "condition" = c("unique", "common"))) }

interactions_unique = lapply(interactions_01, 
                             function(x) find_unique(x, unique))

interactions_unique = ldply(interactions_unique)

```

```{r stack154, fig.width=10, fig.height=8}
p <- ggplot(data = interactions_unique, aes(x = reorder(.id,interactions), y = interactions)) +
  geom_col(aes(fill = condition), width = 0.7)+
  geom_text( data=subset(interactions_unique, interactions!=0),
    position = position_stack(vjust = 0.5), 
    size = 4, 
    aes(y = ,label = interactions, group= condition), color = "white")+

  
    scale_fill_manual(values = c("#00539CFF", "orangered"))+
    theme_classic() + coord_flip() +  
    xlab("Cell line") + ylab("Contact number") + 
    ggtitle("Number of statistically significant interactions with POLR3G (p<0.01)") 
p



```

## Pie chart
```{r pie_chart_prep2, fig.width=4, fig.height=4}
pie_df = data.frame("value" = c(length(unique), length(not_unique)),
                    "condition" = c("unique","common"))
```

```{r piechart2, fig.width=5, fig.height=4}
ggplot(pie_df, aes(x="", y=value, fill=condition)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  scale_fill_manual(values = c("#00539CFF", "orangered")) + 
  geom_text(
    position = position_stack(vjust = 0.5), 
    size = 5, 
    aes(y = ,label = value, group =condition), color = "white")+
    ggtitle("Number of statistically significant interactions \n with POLR3G (p<0.01)") 

```

## BENGI dataset

```{r import_bengi_upset, fig.width=8, fig.height=5}
raw_bengi_polr3g_df = ldply(raw_bengi_polr3g)
interactions = lapply(raw_bengi_polr3g, function(x) x$polr3g_interaction)

upset(fromList(interactions),
      sets.bar.color = "#56B4E9",
      order.by = "freq", #mb.ratio = c(0.4,0.6),
      empty.intersections = NULL, set_size.show = TRUE,
      set_size.scale_max = 30)
```




```{r plot_unique_bengi, fig.width=8, fig.height=5}
# Unique vs common
freq = data.frame(table(raw_bengi_polr3g_df$polr3g_interaction))
unique = as.vector(freq[freq$Freq == 1,]$Var1)
not_unique = as.vector(freq[freq$Freq >= 2,]$Var1)


# counting interactions
find_unique <- function(df, unique){
  tmp = ifelse(df$polr3g_interaction %in% unique, yes = TRUE, no = FALSE)
  return(data.frame("interactions" = c(sum(tmp==TRUE), sum(tmp==FALSE)),
                    "condition" = c("unique", "common"))) }

interactions_unique = lapply(raw_bengi_polr3g, 
                             function(x) find_unique(x, unique))

interactions_unique = ldply(interactions_unique)

p <- ggplot(data = interactions_unique, aes(x = reorder(.id,interactions), y = interactions)) +
  geom_col(aes(fill = condition), width = 0.7)+
  geom_text(data=subset(interactions_unique, interactions!=0),
            position = position_stack(vjust = 0.5),
            size = 4,
            aes(y = , label = interactions, group =condition), color = "white")+
  scale_fill_manual(values = c("#00539CFF", "orangered"))+
  theme_classic() +  
  xlab("Cell line") + ylab("Contact number") + 
  ggtitle("Number of statistically significant interactions with POLR3G (BENGI dataset)") 
p
```





```{r plot_pie_bengi, fig.width=8, fig.height=5}
pie_df = data.frame("value" = c(length(unique), length(not_unique)),
                    "condition" = c("unique","common"))


ggplot(pie_df, aes(x="", y=value, fill=condition)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  scale_fill_manual(values = c("#00539CFF", "orangered")) + 
  geom_text(
    position = position_stack(vjust = 0.5), 
    size = 5, 
    aes(y = ,label = value, group =condition), color = "white")+
  ggtitle("Number of statistically significant interactions \n with POLR3G (BENGI dataset)") 


```




## Capturabilité / nb d'intéractions significative

```{r plot, fig.width=8, fig.height=5}
# Nb interactions / freq moyenne
count_freq = lapply(interactions_05, function(x) x$freq)
count_freq = lapply(count_freq, function(x) 
  return(data.frame("mean_freq" = mean(x), "interaction_number" = length(x))))
count_freq = ldply(count_freq)
row.names(count_freq) = count_freq$.id
count_freq$.id = NULL
```

```{r plot1, fig.width=8, fig.height=5}
ggplot(count_freq, aes(x=mean_freq, y=interaction_number)) +
  geom_point(color = "red") + 
  ggrepel::geom_text_repel(aes(label = rownames(count_freq)),
                 size = 3.5) +
  theme_classic() +
  xlab("Mean Frequency / Capturability") + ylab("Interaction number")

```

```{r test}
# Non normal
shapiro.test(count_freq$mean_freq)

# non significatif
cor.test(~mean_freq+interaction_number, count_freq, method = "spearman")

```







